Simple repression draft

### 2020-06-03 doc/appendix_01_nonbursty.tex
Finish appendix on mean and Fanos for nonbursty models

Next consider the cartoon for nonequilibrium model two in
Figure~\ref{fig1:means_cartoons}. Now we must track probabilities
$p_R$, $p_P$, and $p_E$ for the repressor bound, polymerase
bound, and unbound states, respectively. By analogy
to~\ref{eq:poisson_promoter_cme_appdx}, the master equation for
model two can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and
$k_R^-$, except for RNAP binding and unbinding instead of
repressor. Similarly $p_P(m,t)$ is defined for the active
(RNAP-bound) state exactly as are $p_R(m,t)$ and $p_E(m,t)$ for
the repressor bound and unbound states, respectively.
The new subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are
produced, the promoter state also changes. This is encoded by the
terms involving $r$, the last term in each of the equations for
$p_E$ and $p_P$. The former accounts for arrivals in the
unbound state and the latter accounts for departures from the
RNAP-bound state.

To condense and clarify the unwieldy notation of
Eq.~\ref{eq:model2_cme_appdx}, it can be rewriten in matrix form. We define the column vector $\vec{p}(m,t)$ as
\begin{equation}
\vec{p}(m,t)
= \begin{pmatrix} p_R(m,t) \\ p_E(m,t) \\ p_P(m,t) \end{pmatrix}
\end{equation}
to gather, for a given $m$, the probabilities of finding the
system in the three possible promoter states. Then all the
transition rates may be condensed into matrices which multiply
this vector. The first matrix is
\begin{equation}
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
% \label{eq:3state_cme_matrices_pt1}
\end{equation}
which tracks all promoter state changes in
Eq.~\ref{eq:model2_cme_appdx} that are \textit{not} accompanied
by a change in the mRNA copy number. The two terms accounting for
transcription, the only transition that increases mRNA copy
number, must be handled by two separate matrices given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
% \label{eq:3state_cme_matrices_pt2}
\end{equation}
With these definitions, we can condense Eq.~\ref{eq:model2_cme_appdx} into
the single equation
\begin{equation}
\frac{d}{dt} \vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:generic_cme_appdx}
\end{equation}
which is just Eq.~\ref{eq:3state_rep_cme} in the main text.
Straightforward if tedious algebra verifies that
Eqs.~\ref{eq:3state_rep_cme} and~\ref{eq:generic_cme_appdx} are
in fact equivalent.

@@

\subsubsection{Promoter state probabilities $\langle\vec{m}^0\rangle$}
To begin, we will find the promoter state probabilities
$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_ss} by summing over all
mRNA copy numbers $m$, producing
\begin{equation}
0 = \sum_m \left[
    \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right]
\end{equation}
Using the definitions of $\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$,
and noting the matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$
are all independent of $m$ and can be moved outside the sum, this simplifies to
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_m \vec{p}(m-1)
    + \gamma \sum_m (m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_020}
\end{equation}
The last two terms can be handled by reindexing the summations, transforming
them to match the definitions of $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since
negative numbers of mRNA are nonsensical, we have
\begin{equation}
\sum_{m=0}^\infty \vec{p}(m-1)
= \sum_{m=-1}^\infty \vec{p}(m)
= \sum_{m=0}^\infty \vec{p}(m) = \langle\vec{m}^0\rangle.
\end{equation}
Similarly for the second,
\begin{equation}
\sum_{m=0}^\infty (m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty m\vec{p}(m)
= \sum_{m=0}^\infty m\vec{p}(m) = \langle\vec{m}\rangle,
\end{equation}
which holds since in extending the lower limit from $m=1$ to $m=0$,
the extra term we added to the sum is zero. Substituting these back in
Eq.~\ref{eq:generic_cme_deriv_020}, we have
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \langle\vec{m}^0\rangle
    + \gamma \langle\vec{m}\rangle,
\end{equation}
or simply
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle.
\label{eq:generic_cme_vecm0}
\end{equation}
So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing
a promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this
set of linear equations, subject to the normalization constraint
$\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$.

@@

\subsubsection{Summary of general results}
For ease of reference, we collect and reprint here the key results derived in this section that are used in the main text and subsequent subsections. Mean mRNA copy number and Fano factor are given by Eqs.~\ref{eq:generic_mean_m_appdx} and \ref{eq:generic_fano_appdx}, which are
\begin{equation}
\langle{m}\rangle
= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
\end{equation}
and
\begin{equation}
\nu = 1 - \langle{m}\rangle
+ \frac{\vec{1}^\dagger\cdot \mathbf{R_A}\langle\vec{m}\rangle}
        {\gamma \langle{m}\rangle},
\end{equation}
respectively.
To compute these two quantities, we need the expressions for
$\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$ given by solving
Eqs.~\ref{eq:generic_cme_vecm0} and \ref{eq:generic_cme_deriv_040},
respectively, which are
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle = 0
\end{equation}
and
\begin{equation}
(\mathbf{K} - \mathbf{R_D}
+ \mathbf{R_A} - \gamma\mathbf{I}) \langle\vec{m}\rangle
= - \mathbf{R_A} \langle\vec{m}^0\rangle.
\end{equation}

\subsection{Nonequilibrium Model One - Poisson Promoter}
\mmnote{Derive mean mRNA referred to in means section.}

\subsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
\mmnote{Obtain mean mRNA and Fano factor from CME.}

\subsection{Nonequilibrium Model Three
    - Multistep Transcription Initiation and Escape}
\mmnote{Derive mean mRNA from CME.}
\newline
\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}

\subsection{Nonequilibrium Model Four
    - ``Active'' and ``Inactive'' States}
\mmnote{Derive mean mRNA from CME.}
\newline
\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}

>>>>

Next consider the cartoon for nonequilibrium model two in
Figure~\ref{fig1:means_cartoons}. Now we must track probabilities
$p_R$, $p_P$, and $p_E$ for the repressor bound, empty, and
polymerase bound states, respectively. By analogy
to~\ref{eq:poisson_promoter_cme_appdx}, the master equation for
model two can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and
$k_R^-$, except for RNAP binding and unbinding instead of
repressor. Similarly $p_P(m,t)$ is defined for the active
(RNAP-bound) state exactly as are $p_R(m,t)$ and $p_E(m,t)$ for
the repressor bound and unbound states, respectively.
The new subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are
produced, the promoter state also changes. This is encoded by the
terms involving $r$, the last term in each of the equations for
$p_E$ and $p_P$. The former accounts for arrivals in the
unbound state and the latter accounts for departures from the
RNAP-bound state.

To condense and clarify the unwieldy notation of
Eq.~\ref{eq:model2_cme_appdx}, it can be rewriten in matrix form. We define the column vector $\vec{p}(m,t)$ as
\begin{equation}
\vec{p}(m,t)
= \begin{pmatrix} p_R(m,t) \\ p_E(m,t) \\ p_P(m,t) \end{pmatrix}
\end{equation}
to gather, for a given $m$, the probabilities of finding the
system in the three possible promoter states. Then all the
transition rates may be condensed into matrices which multiply
this vector. The first matrix is
\begin{equation}
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
% \label{eq:3state_cme_matrices_pt1}
\end{equation}
which tracks all promoter state changes in
Eq.~\ref{eq:model2_cme_appdx} that are \textit{not} accompanied
by a change in the mRNA copy number. The two terms accounting for
transcription, the only transition that increases mRNA copy
number, must be handled by two separate matrices given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
% \label{eq:3state_cme_matrices_pt2}
\end{equation}
$\mathbf{R_A}$ accounts for transitions \textit{arriving} in a given state
while $\mathbf{R_D}$ tracks \textit{departing} transitions.
With these definitions, we can condense Eq.~\ref{eq:model2_cme_appdx} into
the single equation
\begin{equation}
\frac{d}{dt} \vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:generic_cme_appdx}
\end{equation}
which is just Eq.~\ref{eq:3state_rep_cme} in the main text.
Straightforward if tedious algebra verifies that
Eqs.~\ref{eq:model2_cme_appdx} and~\ref{eq:generic_cme_appdx} are
in fact equivalent.

@@

\subsubsection{Promoter state probabilities $\langle\vec{m}^0\rangle$}
\label{sec:m0_appdx}
To begin, we will find the promoter state probabilities
$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_ss} by summing over all
mRNA copy numbers $m$, producing
\begin{equation}
0 = \sum_m \left[
    \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1)
\right]
\end{equation}
Using the definitions of $\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$,
and noting the matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$
are all independent of $m$ and can be moved outside the sum, this simplifies to
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_m \vec{p}(m-1)
    + \gamma \sum_m (m+1)\vec{p}(m+1).
\label{eq:generic_cme_deriv_020}
\end{equation}
The last two terms can be handled by reindexing the summations, transforming
them to match the definitions of $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since
negative numbers of mRNA are nonsensical, we have
\begin{equation}
\sum_{m=0}^\infty \vec{p}(m-1)
= \sum_{m=-1}^\infty \vec{p}(m)
= \sum_{m=0}^\infty \vec{p}(m) = \langle\vec{m}^0\rangle.
\end{equation}
Similarly for the second,
\begin{equation}
\sum_{m=0}^\infty (m+1)\vec{p}(m+1)
= \sum_{m=1}^\infty m\vec{p}(m)
= \sum_{m=0}^\infty m\vec{p}(m) = \langle\vec{m}\rangle,
\end{equation}
which holds since in extending the lower limit from $m=1$ to $m=0$,
the extra term we added to the sum is zero. Substituting these back in
Eq.~\ref{eq:generic_cme_deriv_020}, we have
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \langle\vec{m}^0\rangle
    + \gamma \langle\vec{m}\rangle,
\end{equation}
or simply
\begin{equation}
0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle.
\label{eq:generic_cme_vecm0}
\end{equation}
So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing
a promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this
set of linear equations, subject to the normalization constraint
$\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$.
It will turn out to be the case that, if the matrix
$\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}$ is $n$ dimensional, it will always
have only $n-1$ linearly independent equations. Including the normalization
condition provides the $n$-th linearly independent equation, ensuring a
unique solution. So when using this equation to solve for
$\langle\vec{m}^0\rangle$, we may always drop one row of the matrix equation
at our convenience and supplement the system with the normalization
condition instead.

@@

\subsubsection{Summary of general results}
For ease of reference, we collect and reprint here the key results derived
in this section that are used in the main text and subsequent subsections.
Mean mRNA copy number and Fano factor are given by
Eqs.~\ref{eq:generic_mean_m_appdx} and \ref{eq:generic_fano_appdx}, which are
\begin{equation}
\langle{m}\rangle
= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
\label{eq:mean_m_appdx_ref}
\end{equation}
and
\begin{equation}
\nu = 1 - \langle{m}\rangle
+ \frac{\vec{1}^\dagger\cdot \mathbf{R_A}\langle\vec{m}\rangle}
        {\gamma \langle{m}\rangle},
\label{eq:fano_appdx_ref}
\end{equation}
respectively.
To compute these two quantities, we need the expressions for
$\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$ given by solving
Eqs.~\ref{eq:generic_cme_vecm0} and \ref{eq:generic_cme_deriv_040},
respectively, which are
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle = 0
\label{eq:vecm0_appdx_ref}
\end{equation}
and
\begin{equation}
(\mathbf{K} - \mathbf{R_D}
+ \mathbf{R_A} - \gamma\mathbf{I}) \langle\vec{m}\rangle
= - \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:vecm_appdx_ref}
\end{equation}
Some comments are in order before we consider particular models. First,
note that to obtain $\langle\vec{m}\rangle$ and $\nu$, we need not bother
solving for all components of the vectors $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$, but only the components which are multiplied by
nonzero elements of $\mathbf{R_A}$. The only component of
$\langle\vec{m}^0\rangle$ that ever survives is the transciptionally active
state, and for the models we consider here, there is only ever one such state.
This will save us some amount of algebra below.

Also note that we are computing Fano factors to verify the results of
Section~\ref{sec:beyond_means}, concerning the constitutive promoter models
in Figure~\ref{fig2:constit_cartoons} which are analogs of the simple
repression models in Figure~\ref{fig1:means_cartoons}. We can translate
the matrices from the simple repression models to the constitutive case
by simply all occurences of repressor rates and removing the row and column
corresponding to the repressor bound state. The results for
$\langle{m}\rangle$ computed in the repressed case can be easily translated
to the constitutive case, rather than recalculating from scratch, by taking
the limit $k_R^+\rightarrow 0$, since this amounts to sending repressor
copy number to zero.

Finally, we point out that it would be possible to compute
$\langle\vec{m}^0\rangle$ more simply using the diagram methods from King
and Altman~\cite{King1956} (also independently discovered by
Hill~\cite{Hill1966}). But to our knowledge this method cannot be
applied to compute $\langle\vec{m}\rangle$ or $\nu$, so since we would need
to resort to solving the matrix equations anyways for $\langle\vec{m}\rangle$,
we choose not to introduce the extra conceptual burden of the diagram
methods simply for computing $\langle\vec{m}^0\rangle$.

\subsection{Nonequilibrium Model One - Poisson Promoter}
\subsubsection{Mean mRNA}
For nonequilibrium model one in Figure~\ref{fig1:means_cartoons},
we have already shown the full master equation in
Eq.~\ref{eq:poisson_promoter_cme} and Eq.~\ref{eq:poisson_promoter_cme_appdx},
but for completeness we reprint it again as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\end{split}
\end{align}
This is a direct transcription of the states and rates in
Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form
of the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\end{equation}
where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript
and denote both simply by $\mathbf{R}$. Since our interest is only in
steady-state we dropped the time dependence as well.

First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
$p_U$, the probabilities of finding the system in either promoter state,
and note that only $p_U$ survives multiplication by $\mathbf{R}$, since
\begin{equation}
\mathbf{R} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix}
= \begin{pmatrix} 0 \\ r p_U \end{pmatrix},
\label{eq:Rm0_model1_appdx}
\end{equation}
so we need not bother finding $p_R$. Then we have
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle
= \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
\end{equation}
As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
dependent, so taking only the first row and using normalization to set
$p_R = 1-p_U$ gives
\begin{equation}
-k_R^- (1-p_U) + k_R^+ p_U = 0,
\end{equation}
which is easily solved to find
\begin{equation}
p_U = \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
Substituting this into Eq.~\ref{eq:Rm0_model1_appdx}, and the result of that
into Eq.~\ref{eq:mean_m_appdx_ref}, we have
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}
\end{equation}
as asserted in Eq.~\ref{eq:mean_m_model1} of the main text.

\subsubsection{Fano factor}
To verify that the Fano factor for model one in
Figure~\ref{fig2:constit_cartoons} is in fact 1 as claimed in the main text,
note that in this limit $p_U = 1$ and $\langle{m}\rangle = r/\gamma$. All
elements of $\mathbf{K}$ are zero, and $\mathbf{R_A}-\mathbf{R_D} = 0$,
so Eq.~\ref{eq:vecm_appdx_ref} reduces to
\begin{equation}
- \gamma \langle\vec{m}\rangle = - r,
\end{equation}
or, in other words, since there is only one promoter state,
$\langle\vec{m}\rangle = \langle{m}\rangle$. Then it follows that
\begin{equation}
\nu = 1 -\frac{r}{\gamma}
    + \frac{r \langle{m}\rangle}{\gamma \langle{m}\rangle}
= 1
\end{equation}
as claimed.

\subsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
\subsubsection{Mean mRNA}
As shown earlier, the full master equation for model two in
Figure~\ref{fig1:means_cartoons} is
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
    \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U},
\end{split}
\end{align}
which can be condensed to the matrix form of Eq.~\ref{eq:generic_cme_appdx}
with matrices given by
\begin{equation}
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},\
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
\label{eq:model2_matrices_appdx}
\end{equation}
As for model one, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being
found in repressor bound, empty, or RNAP-bound states, respectively.
Only $p_P$ is necessary to find since
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 \\ r p_P \\ 0 \end{pmatrix}.
\label{eq:model2_deriv_appdx_020}
\end{equation}
Then Eq.~\ref{eq:vecm0_appdx_ref} for $\langle\vect{m}\rangle$ reads
\begin{equation}
\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
        k_R^- & -k_R^+ -k_P^+ & k_P^- + r\\
        0 & k_P^+ & -k_P^- - r
\end{pmatrix}
\begin{pmatrix}
    p_R \\ p_E \\ p_P
\end{pmatrix}
= 0.
\label{eq:model2_K-R+R_for_m0}
\end{equation}
Discarding the middle row as redundant and incorporating the normalization
condition leads to a set of three linearly independent equations, namely
\begin{align}
-k_R^- p_R + k_R^+ p_E &= 0 \\
k_P^+ p_E + (-k_P^- - r) p_P &= 0 \\
p_R + p_E + p_P &= 1.
\end{align}
Using $p_R = 1 - p_E - p_P$ to eliminate $p_R$ in the first and solving the
resulting equation for $p_E$ gives
$p_E = (1 - p_P){k_R^-}/{(k_R^- + k_R^+)}$.
Substituting this for $p_E$ in the second equation gives an equation in
$p_P$ alone which is
\begin{equation}
k_P^+ k_R^- (1-p_P) - (k_P^- + r)(k_R^+ + k_R^-) p_P = 0
\end{equation}
and solving for $p_P$ gives
\begin{equation}
p_P = \frac{k_P^+ k_R^-}{k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
\end{equation}
Substituting this in Eq.~\ref{eq:model2_deriv_appdx_020} and multiplying by $\mathbf{R_A}$ produces
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= r \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}
\begin{pmatrix} 0 \\ 1 \\ 0 \end{pmatrix}
\end{equation}
from which $\langle{m}\rangle$ follows readily,
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma}
        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)},
\label{eq:model2_meanm_appdx}
\end{equation}
as claimed in Eq.~\ref{eq:model2_meanm} in the main text.
